A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-06-21 manuscript.qmd
iterate on glmm and simulation-based sections

As study designs become more complex, psychological researchers require more sophisticated statistical models to capture their nuanced relationships and grouping structures [@yarkoniGeneralizabilityCrisis2022]. GLMMs (also called multilevel models) are gaining popularity because they offer great flexibility when applied carefully [@brownIntroductionLinearMixedEffects2021; @matuschekBalancingTypeError2017; @meteyardBestPracticeGuidance2020a].
GLMMs are an extension of LMMs (Linear Mixed Models), which are, in turn, extensions of linear regression models that account for correlated data, including hierarchical structures [@fahrmeirRegressionModelsMethods2021; @bolkerLinearGeneralizedLinear2015]. In this context, correlated data means that the value in the outcome variable for one observation may be related to the value of another observation in a systematic way that is not already accounted for by the usual (fixed) predictor variables (e.g., the age of participants). This correlation can arise for various reasons: For instance, responses to some stimuli from some participants might be more similar because the same person was measured multiple times (repeated measurements), participants belong to the same group (clustering), or participants responded to the same stimulus (stimulus effects). Thus, modeling such correlations is important whenever the data has a clear structure, while the grouping variables can be hierarchically nested (e.g., grouping variables students and schools: each student belongs to exactly one school) or cross-classified (e.g., grouping variables students and math exercises: each student is presented with several math exercises). LMMs are used when the outcome variable is continuous and follows a normal distribution. They allow for the modeling of fixed effects, which capture the relationships between the usual predictors and the outcome, as well as random effects, which account for the different types of correlation structure and grouping effects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across groups. Correlated random effects can be assumed in models that contain both random intercepts and random slopes.
GLMMs extend the LMM framework to accommodate non-normally distributed continuous and categorical outcome variables. GLMMs involve a link function that connects the linear combination of predictor variables to the expected value of the outcome variable. The link function allows for modeling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the outcome variable.

@@

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on formulas to calculate power directly, while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., what percentage of simulated datasets produces a significant p-value). Available formula-based software packages for multilevel models often do not include GLMMs or are limited to simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021; @kainPracticalGuidePower2015; @greenSIMRPackagePower2016a]. However, many of these tutorials focus on linear mixed models (LMMs) and the most common study designs [see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers using more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions: The distribution assumption specifies the distributional family of the outcome variable, after conditioning on all predictor variables. Assumptions about the random effects include the assumption of normality and the covariance structure among the random effects. Making these decisions requires understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in @maxwellSampleSizePlanning2008 and @kumleEstimatingPowerGeneralized2021).

@@

In Table 1, we review existing R packages that can be used for power analysis for GLMMs.

>>>>

As study designs become more complex, psychological researchers require more sophisticated statistical models to capture their nuanced relationships and grouping structures [@yarkoniGeneralizabilityCrisis2022]. GLMMs (also called multilevel models) are gaining popularity because they offer great flexibility when applied carefully [@brownIntroductionLinearMixedEffects2021; @matuschekBalancingTypeError2017; @meteyardBestPracticeGuidance2020a].
GLMMs are an extension of LMMs (Linear Mixed Models), which are, in turn, extensions of linear regression models that account for correlated data, including hierarchical structures [@fahrmeirRegressionModelsMethods2021; @bolkerLinearGeneralizedLinear2015]. In this context, correlated data means that the value in the outcome variable for one observation may be related to the value of another observation in a systematic way that is not already accounted for by the usual (fixed) predictor variables (e.g., the age of participants). This correlation can arise for various reasons: For instance, responses to some stimuli from some participants might be more similar because the same person was measured multiple times (repeated measurements), participants belong to the same group (clustering), or participants responded to the same stimulus (stimulus effects). Thus, modeling such correlations is important whenever the data has a clear structure, while the grouping variables can be hierarchically nested (e.g., grouping variables students and schools: each student belongs to exactly one school) or cross-classified (e.g., grouping variables students and math exercises: each student is presented with several math exercises). LMMs are used when the outcome variable is continuous and follows a normal distribution (after conditioning on all predictor variables). They allow for the modeling of fixed effects, which capture the relationships between the usual predictors and the outcome, as well as random effects, which account for the different types of correlation structure and grouping effects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across groups. Correlated random effects can be assumed in models that contain both random intercepts and random slopes.
GLMMs extend the LMM framework to accommodate non-normally distributed continuous and categorical outcome variables. GLMMs involve a link function that connects the linear combination of predictor variables to the expected value of the outcome variable. The link function allows for modeling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the outcome variable.

@@

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. In Table 1, we review available software packages that can be used to perform power analysis for multilevel models in `r cite_r()`. Power analysis methods can be categorized into formula-based, summary-statistics-based and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on exact formulas to calculate power directly. Summary-statistics-based methods use statistical theory to approximate power based on formula-based methods developed for simple t-tests. Simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically, that is what percentage of simulated datasets produces a significant p-value.
Available formula-based and summary-statistics-based software packages for multilevel models often do not include GLMMs or are limited to simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021; @kainPracticalGuidePower2015; @greenSIMRPackagePower2016a]. However, many of these tutorials focus on LMMs and the most common study designs [see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers using more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions about the model structure that should align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in @maxwellSampleSizePlanning2008 and @kumleEstimatingPowerGeneralized2021). We will discuss a number of strategies on how to specify model parameters for application-specific, tailored data simulations in a later chapter.

@@
